### 2025-05-29 thesis/thesis.tex
Write about relational theory

\newcommand{\figuresource}[1]{
	\begin{center}Quelle: {#1}\end{center}
}

@@

\begin{enumerate}
    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
          training data \citep{, RATSQL}.
    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
          by up to 10.5\% \citep{IRNet, RATSQL}.
    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
\end{enumerate}

@@

The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-SQL}.

@@

>>>>

\newcommand{\figuresource}[1]{
	\begin{center}Source: {#1}\end{center}
}

@@

\begin{enumerate}
    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity
          of approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-
          SQL presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted 
          self attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their 
          transferability. The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over 
          the underlying database schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the 
          \textsc{Spider} benchmark). With standard benchmarks emerging it became easier to verify and quantify which approach had 
          the highest transferability as \textsc{Spider} specifically had independent development and test datasets, preventing 
          approaches from over-optimizing on training data \citep{, RATSQL}.
    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation
          (Seq2SQL) their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized 
          structured decoding, \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges 
          around \textit{schema linking} outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle 
          rephrasings, spelling mistakes and variations in natural language usage far beyond what traditional NL2SQL approaches 
          could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex 
          aggregations, nested structures and joins dramatically improved over the course of the research that happened in this 
          field. Whilst \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural 
          approaches to handle complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation 
          approach introduced by \textsc{IRNet} by up to 10.5\% \citep{IRNet, RATSQL}.
    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to 
          end query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} 
          showed novel and state-of-the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to 
          accurately reason about user intent and traverse the database schema while generating queries \citep{TypeSQL, GNN, 
          RATSQL}. 
\end{enumerate}

@@

The comparison between zero-shot and few-shot performance reveals the accuracy gain potential through supplying examples
to models in the inference context. While contemporary LLMs (such as GPT-4) have demonstrate impressive zero-shot performance
(achieving 72.3\% execution accuarcy on benchmarks like \textsc{Spider}) \citep[Table 1, p.~8]{DAIL-SQL}, few-shot learning 
still shows to substantially improve model performance. \cite{DAIL-SQL} shows that even one-shot learning boosts GPT-4's
execution accuracy to 80.2\%, representing a 7.9\% increase, while five-shot learning reaches 82.4\% \citep[Table 2, p.~8]{DAIL-
SQL}.

@@

This section serves as an introduction to the theoretical concepts utilized in this thesis. It is meant to establish a
baseline of understanding for the design, implementation and evaluation phases. This section briefly covers relational
database theory (incl. SQL), machine learning fundamentals and their subsequent application for large language models
and the inner workings of PostgreSQL and its extension architecture at a high level.

\subsection{Relational Database Theory}

The relational database theory provides a mathematical framework for formalizing data layout in database management
systems (DBMS), enabling efficient storage and retrieval mechanisms. This theory is helps to understand the way in
which natural language queries need to be translated into database operations and what challenges arise during this
translation.

\subsubsection{Relational Model Fundamentals}

The relational model was introduced by \citeauthor{RelationalModel} in \citeyear{RelationalModel} and essentially forms
the theoretical foundation for most modern database systems \citep{RelationalModel, Rendezvous}. A relational database
organizes data in relations (ie. tables), where each relation consists of tuples (ie. rows) containing attributes
(ie. columns) of specific domains (ie. data types and constraints). The mathematical origins of this model root in
set theory and first-order predicate logic \citep{RelationalModel, DatabaseSystemIntro}.

Furthermore the relational model abstracts data storage and respective physical implementation details by design,
allowing interactions with data through respective logical structures (relations, tuples, attributes etc.) rather than
exposing the underlying storage mechanism. This abstraction is fundamental for NL2SQL systems, as natural language
queries can be mapped into queries on these structures regardless of underlying physical storage implementation,
allowing for better transferability of NL2SQL systems (e.g. across different database deployments).

\subsubsection{Core Concepts}

The core concepts of the relational model that are relevant for this thesis include:

\begin{itemize}
    \item \textbf{Relations} — Mathematical sets of tuples representing entities (e.g. employees, products) or
          relationships (e.g. enrollment, purchases). Each relation has a fixed structure.
    \item \textbf{Tuples} — Tuples represent rows of data, by grouping a several attributes into one logical unit
          (e.g. one specific purchase, containing the amount, date and shop). All tuples in the same relation share
          the same structure.
    \item \textbf{Attributes} — Attributes are frequently referred to as columns, representing a specific dimension
          of a tuple (e.g. customers have names). Attributes are represented in a fixed domain.
    \item \textbf{Domains} — Domains are sets of allowed values for attributes, including mathematically formalizing
          data types and constraints.
    \item \textbf{Schemas} — Structural definitions specifying relations, their attribute names, domains, and integrity
          constraints.
    \item \textbf{Primary Keys} — Attributes that uniquely identify tuples within a relation. These
          ensure entity integrity and ensure referential relationships.
    \item \textbf{Foreign Keys} — Attributes referencing primary keys in other relations, allowing databases to maintain
          referential integrity and enabling semantically correct complex queries across multiple tables through joins.
\end{itemize}

These concepts form the semantic foundation that natural language interfaces must navigate when translating user queries
into formal database operations. As shown in the literature review \textit{schema-linking} is a crucial problem of
NLIDBs — establishing an understanding of which relations a natural language query refers to is the first step of
formalizing a database operation for retrieval.

\subsubsection{SQL as a Declarative Query Language}

Structured Query Language (SQL) is a defacto standard for interfacing with relational databases. It is a declarative
language to describe database operations like selection, insertion, updating and deletion of database contents. SQL's
design is strongly influenced by relational algebra and tuple relational calculus. It enables complex data retrieval
through a readable query syntax incorporating relational algebra operations like:

\begin{itemize}
    \item \textbf{Selection} ($\sigma$) — Filtering tuples based on specified conditions
    \item \textbf{Projection} ($\pi$) — Extracting specific attributes from relations
    \item \textbf{Joins} ($\bowtie$) — Combining data from multiple relations based on common attributes
    \item \textbf{Aggregation} — Computing summary statistics over groups of tuples
\end{itemize}

The basic syntax of SQL queries follows a logical pattern that reflects these theoretical operations:

\begin{verbatim}
SELECT attributes      -- Project attributes
FROM relations         -- Specify relations to use
WHERE conditions       -- Selection of tuples
GROUP BY attributes    -- Grouping for aggregation
HAVING conditions      -- Group-level filtering
ORDER BY attributes    -- Ordering of results
\end{verbatim}

There are further operations like various types of \texttt{JOIN}s, \texttt{OFFSET}, \texttt{LIMIT}, \texttt{WITH}
that are heavily used in SQL which are excluded for readability and simplicity.

The nature of SQL creates a fundamental challenge for NL2SQL systems; natural language queries express (often fuzzy) user
intent in terms of desired output (e.g., ``Give me 5 great movies'') while SQL formalizes an explicit and discrete way
of how to retrieve this information from the database:

\begin{verbatim}
SELECT m.title, AVG(r.score) as rating
FROM movies m
JOIN ratings r ON m.movie_id = r.movie_id
GROUP BY m.movie_id, m.title
HAVING AVG(r.score) >= 8
ORDER BY rating DESC
LIMIT 5
\end{verbatim}

This semantic gap between natural language and formal queries represents the core challenge that this thesis addresses.

Furthermore, SQL's compositional nature allows complex queries to be built from simpler components through nesting and
combination of operations. NL2SQL systems must therefore not only fill in the individual query components but also how
to compose these to semantically accurate and syntactically correct query statements. This is particularly important when
dealing with complex natural language queries that may require multi-clause SQL statements involving subqueries, multiple
joins, and aggregation functions to answer.

\subsubsection{Normalization and Schema Design}

Database schemas typically follow normalization principles to eliminate redundancy and maintain data integrity. The normal forms
(1NF, 2NF, 3NF, BCNF) provide a set of design guidelines and rules for relational database design to reduce duplication,
inconsistencies and anomalies \citep{DatabaseSystemIntro}. Understanding normalization forms is crucial for NL2SQL systems to handle:

\begin{itemize}
    \item \textbf{Schema complexity} — Normalized schemas often distribute logically related information across multiple tables,
          requiring natural language interfaces to understand implicit relationships and generate appropriate joins.
    \item \textbf{Semantic mapping} — Humans typically think about data in denormalized, conceptual terms, while relational databases
          store data in its normalized form. NL2SQL systems must overcome this layout difference (e.g. there is no 1:1 mapping
          of concepts to tables).
    \item \textbf{Query complexity} — Retrieving simple information from relational databases may require multiple joins in normalized schemas,
          challenging NL2SQL systems to generate potentially complex SQL statements from simple user requests.
\end{itemize}

The tension between normalized database design and intuitive natural language use represents a key challenge that influences
the architecture and design decisions explored in subsequent sections of this thesis.

\subsection{Machine Learning Fundamentals}
